``Interestingness'' in Mathematical Theory Exploration

### 2019-04-26 PhDThesis/background_theoryexploration.tex
Import a bunch of text and fiddle

@@

We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
Haskell code (these may be fed into another tool, such as \hspec{}, for
proving). These conjectures are arrived at through the following stages:

\iffalse TODO: Make this more formal?
 V \in Var
 F \in Fun
 T \in Term
 T ::= V | F | T1 T2

 Term ::= VAR | Const | Fun (Term)
or
 Term t ::= x | f | t t'
\fi

@@

Such $conjectures$ can be used in several ways: they can be simplified for
direct presentation to the user (by removing any equation which can be derived
from the others by rewriting), sent to a more rigorous system like \hspec{} or
\textsc{Hipster} for proving, or even serve as a background theory for an
automated theorem prover \cite{claessen2013automating}.

>>>>

% TODO Work this in
\iffalse
Theory exploration is the task of taking a \emph{signature} of definitions in
some formal system (for example a programming language) and automatically
generating a set of formal statements (properties) involving those
definitions. These may be conjectures or theorems (proven either by sending
conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps); in either case, these statements
must also be ``interesting'' in some way. It is beyond the scope of this paper
to define what makes a mathematical statement ``interesting'', but this problem
has been tackled extensively in the literature~\cite{colton2000notion}.
\fi
% END TOTO

@@

We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about Haskell
code (these may be fed into another tool, such as \hspec{}, for proving). We
have used version 1 of \qspec{} in our experiments, due to its availability,
stability and tooling integration; hence references to \qspec{} are to version 1
unless stated otherwise. At the time of writing there is a \qspec{} version 2
available, which has a much improved generation procedure which can be
significantly faster than its predecessor. Whilst \qspec{} 2 has advanced the
state of the art in theory exploration, our preliminary experience has found
that it still suffers the scaling issues we identify in this work; indeed, we
even encountered memory exhaustion from some of the examples included with the
\qspec{} 2 source code! Since \qspec{} 2 has less integration with the tooling
we have used, we leave a more thorough analysis of it (and related systems such
as \speculate{}) for future work, pending the necessary infrastructure changes
that would require.

\begin{figure}
  \centering
  \begin{minted}{haskell}
    -- A datatype with two constructors (Note that -- introduces a comment)
    data Bool = True | False

    -- A recursive datatype (S requires a Nat as argument)
    data Nat = Z | S Nat

    -- A function turning a Bool into a Bool
    not :: Bool -> Bool
    not True  = False
    not False = True

    -- A pair of mutually recursive functions

    odd :: Nat -> Bool
    odd    Z  = False
    odd (S n) = even n

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{minted}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some
    simple function definitions (with type annotations).}
  \label{fig:haskellexample}
\end{figure}

\qspec{} (version 1) is written in the Haskell programming language, and works
with definitions which are also written in Haskell. A thorough description of
Haskell, including its suitability for theory exploration, can be found
in~\ref{sec:haskell}. For illustrative purposes, some simple Haskell definitions
are shown in Figure~\ref{fig:haskellexample}. The following procedure is used to
generate conjectures, which are both plausible (due to testing on many examples)
and potentially interesting (due to being mutually irreducible):

@@

\iffalse
% TODO: Compare the above to this; merge/swap anything that's better
\begin{enumerate}
\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
  generates a list $terms$ containing the constants (including functions) from
  $\Sigma$, the variables from $V$ and type-correct function applications
  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
  is finite, function applications are only nested up to a specified depth (by
  default, 3).
\item The elements of $terms$ are grouped into equivalence classes, based on
  their type.
\item The equivalence of terms in each class is tested using \qcheck{}:
  variables are instantiated to particular values, generated randomly, and the
  resulting closed expressions are evaluated and compared for equality.
\item If a class is found to have non-equal members, it is split up to separate
  those members.
\item The previous steps of testing and splitting are repeated until the classes
  stabilise (i.e. no differences have been observed for some specified number of
  repetitions).
\item For each class, one member is selected and equations are conjectured that
  it is equal to each of the other members.
\item A congruence closure algorithm is applied to these equations, to discard
  any which are implied by the others.
\end{enumerate}
\fi

Such conjectures can be used in several ways: they can be simplified for direct
presentation to the user (for example via the congruence closure algorithm
\iffalse\cite{TODO}\fi), sent to a more rigorous system like \hspec{} or
\hipster{} for proving, or even serve as a background theory for an
automated theorem prover \cite{claessen2013automating}.

### 2019-04-26 PhDThesis/bucketing.tex
Import a bunch of text and fiddle

\section{Introduction}

\emph{(Mathematical) Theory Exploration} (TE) describes an open-ended search for
interesting properties (proven or merely conjectured, depending on the approach)
of a given set of formal definitions, such as a mathematical theory, a
scientific model or a software library. It has been applied to generating
``background lemmas'' for the interactive proof assistant
\textsc{Isabelle/HOL}~\cite{johansson2009isacosy,Hipster} and automated
theorem provers~\cite{claessen2013automating}, as well as for generating
specifications for (pure functional) programming
libraries~\cite{QuickSpec,smallbone2017quick,braquehais2017speculate}.

Despite the clever search strategies employed by these tools, they all rely on
enumerating combinations of definitions in some way, which can cause infeasible
running times on larger inputs (in our experience, a few dozen definitions). To
avoid this, users of these systems must carefully select only small subsets of
their definitions to explore at a time.

We dub such selection the \emph{bucketing problem} for theory exploration and
analyse its effect on the running time of theory exploration tools on a
particular problem set, and on the quality of the statements they are able to
generate. An automated solution to this bucketing problem would accept a large
number of definitions and efficiently select sub-sets which are both small
enough to reasonably explore, whilst still enabling interesting properties to be
discovered.

Our contributions are:

\begin{enumerate}
  \item Defining the \emph{bucketing problem} for theory exploration.
  \item Statistical analysis of theory exploration performance on a large
    problem set, derived from existing theorem proving problem sets.
  \item Empirical analysis of the effect of bucketing on the quantity of
    desirable properties attainable by theory exploration.
  \item Experimental application of machine learning to the bucketing problem,
    via unsupervised \emph{clustering}.
  % \item A novel feature extraction method for transforming Haskell expressions
  %   into a form amenable to off-the-shelf learning algorithms.
  % \item An implementation of these feature extraction and theory exploration
  %   approaches.
  % \item A comparison of our methods with existing approaches, both for theory
  %   exploration in Haskell, and for machine learning in other languages.
\end{enumerate}

We describe theory exploration in more detail, as well as the \qspec{} tool we
use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
bucketing problem more formally in \S~\ref{sec:bucketing} and measure its
effect on theory exploration systems in \S~\ref{sec:implementation}. In
\S~\ref{sec:clustering} we highlight the difference betwee bucketing and the
well-known \emph{clustering problem} from machine learning, by demonstrating
that clustering algorithms perform poorly at bucketing.  A variety of related
work is surveyed in \S~\ref{sec:related} and we conclude in
\S~\ref{sec:conclusion} with potential directions for future research.

\section{Background}
\label{sec:background}

\subsection{Theory Exploration}
\label{sec:theoryexploration}

Theory exploration is the task of taking a \emph{signature} of definitions in
some formal system (for example a programming language) and automatically
generating a set of formal statements (properties) involving those
definitions. These may be conjectures or theorems (proven either by sending
conjectures to an automated theorem prover, or by having the generating
procedure proceed in logically sound steps); in either case, these statements
must also be ``interesting'' in some way. It is beyond the scope of this paper
to define what makes a mathematical statement ``interesting'', but this problem
has been tackled extensively in the literature~\cite{colton2000notion}.

We will follow the tradition of prior practitioners and use an existing
corpus of theorems as a \emph{ground truth} against which to compare results.
Existing work has used libraries of the \textsc{Isabelle/HOL} proof assistant as
a ground
truth~\cite{Montano-Rivas.McCasland.Dixon.ea:2012,johansson2009isacosy,claessen2013automating},
but these only contained a few definitions and theorems each. To analyse
behaviour on a larger scale, as required to study the effects of bucketing, we
have chosen the \emph{Tons of Inductive Problems} (TIP) problem set for
automated theorem provers~\cite{claessen2015tip}.  Reasons for choosing TIP
include:

\begin{itemize}
\item Each problem in TIP clearly distinguishes between the statement to be
  proved and the definitions involved, such that the two can be easily and
  meaningfully separated to act as inputs and desired outputs in a TE setting.
  This would not be the case for problem sets like
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item The TIP problem set includes a tool to translate from the language it
  uses to define problems (a variant of SMTLIB) into various other languages
  including Haskell and Isabelle, for which TE tools already exist.
\item TIP problems are relevant to the domains we are interested in, including
  pure functional programming, higher-order functions and inductively defined
  datatypes. This makes TIP more attractive than problem sets for first-order
  languages/logics such as TPTP~\cite{sutcliffe2009tptp}.
\item Many of TIP's problem statements re-use the same definitions, which makes
  it more useful as a ground-truth than a mixture of unrelated problems. This is
  because non-membership in the ground truth will be treated as being
  ``uninteresting''.
\end{itemize}

All existing theory exploration systems rely on brute force enumeration to some
degree. We focus on \qspec{}~\cite{QuickSpec}, which conjectures equations about
Haskell code. These may be fed into another tool, such as \hspec{}, for
subsequent proving. We have used version 1 of \qspec{} in our experiments, due
to its availability, stability and tooling integration, and we will drop the
version number from now on. At the time of writing there is a \qspec{} version 2
available, which has a much improved generation procedure that is significantly
faster than its predecessor. However, it still scales poorly in our experience,
and has less integration with the tooling we have used. We hence leave analysis
of \qspec{} 2 as future work, pending the necessary infrastructure changes that
would require.

\qspec{} (version 1) is written in the Haskell programming language, and works
with definitions which are also written in Haskell. Some simple Haskell
definitions are shown in Figure~\ref{fig:haskellexample}. The following
procedure is used to generate conjectures, which are both plausible (due to
testing on many examples) and potentially interesting (due to being mutually
irreducible):

\begin{figure}
  \centering
  \begin{minted}{haskell}
    -- A datatype with two constructors
    data Bool = True | False

    -- A recursive datatype (S requires a Nat as argument)
    data Nat = Z | S Nat

    -- A function turning a Bool into a Bool
    not :: Bool -> Bool
    not True  = False
    not False = True

    -- A pair of mutually recursive functions

    odd :: Nat -> Bool
    odd    Z  = False
    odd (S n) = even n

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{minted}
  \caption{Haskell datatypes for booleans and natural numbers, followed by some
    simple function definitions (with type annotations).}
  \label{fig:haskellexample}
\end{figure}

\begin{enumerate}
\item Given a typed signature $\Sigma$ and set of variables $V$, \qspec{}
  generates a list $terms$ containing the constants (including functions) from
  $\Sigma$, the variables from $V$ and type-correct function applications
  \hs{f x}, where \hs{f} and \hs{x} are elements of $terms$. To ensure the list
  is finite, function applications are only nested up to a specified depth (by
  default, 3).
\item The elements of $terms$ are grouped into equivalence classes, based on
  their type.
\item The equivalence of terms in each class is tested using \qcheck{}:
  variables are instantiated to particular values, generated randomly, and the
  resulting closed expressions are evaluated and compared for equality.
\item If a class is found to have non-equal members, it is split up to separate
  those members.
\item The previous steps of testing and splitting are repeated until the classes
  stabilise (i.e. no differences have been observed for some specified number of
  repetitions).
\item For each class, one member is selected and equations are conjectured that
  it is equal to each of the other members.
\item A congruence closure algorithm is applied to these equations, to discard
  any which are implied by the others.
\end{enumerate}

@@

\subsection{The Bucketing Problem}
\label{sec:bucketing}

@@

\iffalse
As an example of this recurrent process, we can consider the Peano arithmetic functions from Figure \ref{fig:coreexample}. A valid topological ordering is given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding Core expressions to save space):

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$). A more extensive investigation of these features requires a
concrete implementation, in particular to pin down values for the parameters
such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.  \fi

>>>>

\section{The Signature Selection Problem}

Despite the clever search strategies employed by theory exploration tools, they
all rely on enumerating combinations of definitions in some way, which can cause
infeasible running times on larger inputs (in our experience, a few dozen
definitions). To avoid this, users of these systems must carefully select only
small subsets of their definitions to explore at a time.

We dub such selection the \emph{signature selection problem} for theory
exploration, and we analyse its effect on the running time of theory exploration
tools on the TIP particular problem set, and on the quality of the statements
they are able to generate. An automated solution to this signature selection problem would
accept a large number of definitions and efficiently select sub-sets which are
both small enough to reasonably explore, whilst still enabling interesting
properties to be discovered.

In this chapter we will:

\begin{enumerate}
  \item Define the \emph{signature selection problem} for theory exploration.
  \item Analyse the performance of theory exploration on the TEBenchmark problem
    set.
  \item Analyse the effect of signature selection on the quantity of desirable properties
    attainable by theory exploration on this problem set.
  \item Demonstrate how the signature selection problem is distinct from the well-know
    \emph{clustering problem} in machine learning, and how clustering algorithms
    cannot be directly applied to signature selection.
\end{enumerate}

Some bespoke components required for these analyses and comparisons may also be
useful on their own, or re-purposed for other tasks. These include:

\begin{enumerate}
\item A novel feature extraction method for transforming Haskell expressions
  into a form amenable to off-the-shelf machine-learning algorithms, based on
  the existing \emph{recurrent clustering} algorithm from other languages.
\item An executable implementation of this feature extraction
  approach\footnote{https://github.com/warbo/ml4hsfe}.
\item End-to-end automation for exploration arbitrary, user-provided Haskell
  packages\footnote{https://github.com/warbo/haskell-te}.
\item Dynamic evaluation of Haskell code, with access to dynamically installed
  Haskell packages\footnote{http://hackage.haskell.org/package/nix-eval}.
\item Translation of the signature selection problem to the constrain-solving domain, and
  executable oracles for optimal (and pessimal) signature selection, when the desired
  properties are already known\footnote{Part of
    https://github.com/warbo/bucketing-algorithms}.
\end{enumerate}

\iffalse
% TODO: Do we need this?
We describe theory exploration in more detail, as well as the \qspec{} tool we
use in our experiments, in \S~\ref{sec:theoryexploration}. We introduce the
signature selection problem more formally in \S~\ref{sec:sigselect} and measure its
effect on theory exploration systems in \S~\ref{sec:implementation}. In
\S~\ref{sec:clustering} we highlight the difference between signature selection and the
well-known \emph{clustering problem} from machine learning, by demonstrating
that clustering algorithms perform poorly at signature selection.  A variety of related
work is surveyed in \S~\ref{sec:related} and we conclude in
\S~\ref{sec:conclusion} with potential directions for future research.
\fi

\subsection{Background}
\label{sec:background}

@@

\subsection{The Signature Selection Problem}
\label{sec:sigselect}

@@

\iffalse

As an example of this recurrent process, we can consider the Peano arithmetic
functions from Figure \ref{fig:coreexample}. A valid topological ordering is
given in Figure \ref{fig:sccexample}, which can be our value for $d$ (eliding
Core expressions to save space):

@@

Even in this very simple example we can see a few features of our algorithm
emerge. For example, \hs{odd} and \hs{even} will always appear in the same
cluster, since they only differ in their choice of constructor names (which are
discarded by $toTree$) and recursive calls (which are replaced by
$f_{recursion}$). A more extensive investigation of these features requires a
concrete implementation, in particular to pin down values for the parameters
such as $r$, $c$, $f_{recursion}$, $\alpha$ and so on.

\subsection{Experiments}

% TODO: This was taken from the Background section, so probably doesn't fit here
% without some fiddling.
% TODO: Describe all of our experimental setups:
%
%  - The use of the TEBenchmark methodology for running QuickSpec, including the
%    sampling process
%
%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
%    including how we picked the sizes
%
%  - The definition of HashSpec and its use as control; how we ran all of the
%    same samples through it
%
%  - The use of optimal and pessimal oracles, defined via constraint
%    satisfaction. How we it was only feasible to run these up to size 10 (11?)

We follow the tradition of prior practitioners and use an existing corpus of
theorems as a \emph{ground truth} against which to compare results. To analyse
behaviour on large inputs (i.e. those for which exploration is currently
infeasible), as required to study the effects of signature selection, we have used the
\tebenchmark{} of section~\ref{sec:tebenchmark}, which is based on the Tons of
Inductive Problems (TIP) problem set for theorem provers~\cite{claessen2015tip}.

### 2019-04-26 PhDThesis/intro.tex
Import a bunch of text and fiddle

We begin our undertaking in \S \ref{sec:background} by introducing the Haskell
language, as well as the relevant fields of verification for context. We define
a formal framework for our investigation, and show how it relates to the
existing theorem proving landscape. A selection of theorem proving scenarios
which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst
related work, including existing defintions of interestingness, is surveyed in
\S \ref{sec:related}. We also review the use of exploration in other fields of
Artificial Intelligence and Machine Learning, where researchers are
experimenting with replacing \emph{explicit} goals and rewards with
\emph{implicit} alternatives such as interestingness. Recent efforts in this
area have lead to the emergence of principled theories, mostly based around
(algorithmic) information theory, which may be adapted to our theory exploration
context.

We discuss our present contributions in \S \ref{sec:current} and future research
directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.

>>>>

\iffalse
We begin our undertaking in \S \ref{sec:background} by introducing the Haskell
language, as well as the relevant fields of verification for context. We define
a formal framework for our investigation, and show how it relates to the
existing theorem proving landscape. A selection of theorem proving scenarios
which \emph{require} exploration are discussed in \S \ref{sec:examples}, whilst
related work, including existing defintions of interestingness, is surveyed in
\S \ref{sec:related}. We also review the use of exploration in other fields of
Artificial Intelligence and Machine Learning, where researchers are
experimenting with replacing \emph{explicit} goals and rewards with
\emph{implicit} alternatives such as interestingness. Recent efforts in this
area have lead to the emergence of principled theories, mostly based around
(algorithmic) information theory, which may be adapted to our theory exploration
context.

% TODO: Do we need this?
We discuss our present contributions in \S \ref{sec:current} and future research
directions in \S \ref{sec:future}, before concluding in \S \ref{sec:conclusion}.
\fi

### 2019-04-26 PhDThesis/litreview_staticanalysis.tex
Import a bunch of text and fiddle

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\emph{QuickCheck} does this randomly, \emph{SmallCheck}
enumerates values, etc.).

>>>>

We can reduce the amount of unknown parameters in a concrete program by
propagating statically-known values, for example using
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}; however this can
dramatically increase code size, as each function may be duplicated into
multiple specialised instances. Related techniques, like \emph{symbolic
  execution} (a form of \emph{abstract interpretation}), allow approximate
execution of code, where unknown runtime data is represented as opaque
metavariables. Property- and model-checkers are similar, except they work
empirically by performing multiple runs, instantiating metavariables with
synthetic data (\qcheck{} does this randomly, \textsc{SmallCheck} enumerates
values, etc.).

### 2019-04-26 PhDThesis/litreview_testing.tex
Import a bunch of text and fiddle

Automated testing is a common method of software quality assurance. In Haskell,
some of the most popular testing frameworks go beyond the industry standard of
\emph{unit testing} (asserting that components of a system behave as expected on
a fixed set of inputs) to instead perform \emph{property checking}. A
``property'', in this sense, is a statement about a program's components, which
may include free variables. Such properties are typically encoded in Haskell as
functions, accepting values for the free variables as arguments, and returning a
boolean value (perhaps adorned with useful metadata). A property checker
proceeds by
\begin{enumerate*}
\item generating suitable values for the function's arguments,
\item calling the function with these values,
\item inspecting the return value and
\item repeating until some stop condition is satisfied
\end{enumerate*}.

Most frameworks will test each property a fixed number of times (e.g. 100), and
stop early if the function returns a false value. The major difference between
implementations is the way they instantiate free variables (or, in encoded form,
how they generate arguments for the property functions).

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \texttt{forAll}), or by
choosing a default generator for types implementing its \texttt{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way).

Random testing is effective at thoroughly exercising the branches of a program,
since inputs may be quite large and cover many cases, especially for functions
which recurse on each \emph{sub-}expressions of their input as well. The output
of a random generator will also vary across runs, \emph{increasing} our
confidence each time we check a property, rather than simply reiterating past
results. However, there are two common problems with random property checking.
Firstly we must be able to generate random values, which can be tricky for
recursive datatypes. Consider a type of binary trees, with constructors
\texttt{Leaf~::~Tree} and \texttt{Node~::~Tree~->~Tree~->~Tree}. We might
generate a \texttt{Tree} by picking uniformly between these two constructors
and, if a \texttt{Node} was picked, recursing to generate the two
sub\texttt{Tree}s. Yet this simple scheme will generate \texttt{Tree}s whose
expected size is \emph{unbounded}! This is due to the 50\% chance of generating
a \texttt{Node}, which requires two recursive calls; hence we can expect each
call to recurse $0.5~\times~2~=~1$ more time on average, and that call will also
recurse once more on average, and so on; this can easily exhaust all available
memory. Many automated data generators, such as those from Haskell's
\texttt{derive} package, suffer from this problem. It can be avoided by biasing
the generators (in the \texttt{Tree} example we would produce a \texttt{Leaf}
more than half of the time), but this bias compounds at each level of recursion,
limiting testing to very small values. More sophisticated solutions exist, such
as the \texttt{boltzmann-sampling} package (see \cite{duchon2004boltzmann}) and
even domain-specific languages~\cite{lampropoulos2017beginner}, but all of these
add complications to the property checking process.

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \texttt{shrink} function along with their
generator, which is given any counterexample found during testing and returns a
list of smaller values. In the above example we might shrink a \texttt{Node} by
returning each sub-tree, and return an empty list when asked to shrink a
\texttt{Leaf}. After a counterexample has been found, the same property is
applied to each of the shrunk values in an attempt to narrow down the source of
the problem. When a value cannot be shrunk any more, or if none of its shrunken
versions are counterexamples, that value is presented to the user in the hope
that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumerating function or using its
\texttt{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package). In the \texttt{Tree} example this might
check \texttt{Leaf}, then \texttt{Node Leaf Leaf}, then
\texttt{Node (Node Leaf Leaf) Leaf}, and so on. This ensures that
counterexamples are always small, that tests are not repeated with the same
values and that memory is not exhausted (since the stopping condition will
trigger long before the enumeration reaches such large values). The disadvantage
is that large inputs (which exercise properties, especially recursive ones, the
most) may occur so late in the enumeration that testing never reaches them.

>>>>

Automated testing is a common method of software quality assurance. Typical
industry standards (outside of specialised domains like those which are
life-critical) include \emph{unit testing} \iffalse \cite{TODO} \fi: asserting
that components of a system behave as expected on a fixed set of inputs. In
Haskell, some of the most popular testing frameworks go beyond this to instead
perform \emph{property checking}. A ``property'', in this sense, is a statement
about a program's components, which may include free variables. We can think of
a unit test as proving an \emph{existential} property: that the given statement
holds for at least one assignment of the variables, and those values are stated
as part of the unit test. However, the phrase ``property checking'' is typically
reserved for \emph{universal} properties: those which should hold \emph{for all}
values of the variables.

Universal properties are typically encoded in Haskell as \emph{functions},
accepting values for the free variables as arguments, and returning a boolean
value (perhaps adorned with useful metadata). If the type of the variables is
finite (for example booleans), it might be feasible to check all possible
values. Typically this is not feasible, especially when multiple variables are
involved, or the input types contain an infinite number of values (for example
lists). Property checkers avoid this by only testing a sub-set of values, in the
hope that a counterexample will be found (if one exists). They work by:

\begin{enumerate}
\item Generating suitable values for the function's arguments
\item Calling the function with these values
\item Inspecting the return value
\item Repeating until some stop condition is satisfied
\end{enumerate}.

Most frameworks will test each property a fixed number of times (e.g. 100), and
stop early if the function returns a false value. The major difference between
implementations is the way they generate values to test with.

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \texttt{forAll}), or by
choosing a default generator for types implementing its \texttt{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
\cite{TODO} \fi).

Random testing is effective at thoroughly exercising the branches of a program,
since inputs may be quite large and cover many cases, especially for functions
which recurse on each \emph{sub-}expressions of their input as well. The output
of a random generator will also vary across runs, rather than checking the same
values every time. Hence every invocation of a property checker, e.g. as part of
a development or debugging workflow, allows more of the input space to be
tested. However, there are two common problems with random property checking.

Firstly we must be able to generate random values, which can be tricky for
recursive datatypes. Consider a type of binary trees, with constructors
\texttt{Leaf~::~Tree} and \texttt{Node~::~Tree~->~Tree~->~Tree}. We might
generate a \texttt{Tree} by picking uniformly between these two constructors
and, if a \texttt{Node} was picked, recursing to generate the two
sub\texttt{Tree}s. Yet this simple scheme will generate \texttt{Tree}s whose
expected size is \emph{unbounded}! This is due to the 50\% chance of generating
a \texttt{Node}, which requires two recursive calls; hence we can expect each
call to recurse $0.5~\times~2~=~1$ more time on average, and that call will also
recurse once more on average, and so on; this can easily exhaust all available
memory. Many automated data generators, such as those from Haskell's
\texttt{derive} package, suffer from this problem. It can be avoided by biasing
the generators (in the \texttt{Tree} example we would produce a \texttt{Leaf}
more than half of the time), but this bias compounds at each level of recursion,
limiting testing to very small values. More sophisticated solutions exist, such
as the \texttt{boltzmann-sampling} package (see \cite{duchon2004boltzmann}) and
even domain-specific languages~\cite{lampropoulos2017beginner}, but all of these
add complications to the property checking process.

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \texttt{shrink} function along with their
generator, which is given any counterexample found during testing and returns a
list of smaller values. In the above example we might shrink a \texttt{Node} by
returning each sub-tree, and return an empty list when asked to shrink a
\texttt{Leaf}. If one of these smaller values is also a counterexample, it too
is shrunk and the process repeats, otherwise it is discarded and the next shrunk
value is tried. When a value cannot be shrunk any more, or if none of its
shrunken versions are counterexamples, that value is presented to the user in
the hope that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumeration function or using its
\texttt{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
\texttt{Tree} example this might check \texttt{Leaf}, then \texttt{Node Leaf
  Leaf}, then \texttt{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
order of ``size'' ensures that counterexamples are always small, that tests are
not repeated with the same values and that memory is not exhausted (since the
stopping condition will trigger long before the enumeration reaches such large
values). The disadvantage is that large inputs (which exercise properties,
especially recursive ones, the most) may occur so late in the enumeration that
testing never reaches them.
